\setchapterpreamble[u]{\margintoc}
\chapter{Conclusions \& Perspectives}
\labch{conclusion}

In this thesis, numerical investigations pertaining to two different aspects 
of liquid-gas interfacial flows were carried out using advanced Volume-of-Fluid methods. 
The first involves understanding and eventually suppressing certain types of 
numerical instabilities that arise when dealing with flows having marked contrasts in density,
particularly in the context of low to moderate numerical resolution. 
The principle of maintaining consistency between the discrete transport of
the discontinuous mass and momentum fields were implemented  
using a conservative discretization of the two phase 
Navier-Stokes equations on staggered Cartesian grids. 
The \textit{shifted fractions} and \textit{sub-grid} methods were developed
in order to address the challenge of consistent advection on staggered grids,
coupled with the extension of the conservative direction-split mass transport
algorithm of Weymouth and Yue \cite{wy} to the direction-split transport of momentum.
Several performance aspects of the aforementioned methods were presented using 
standard benchmark cases such as the behavior of spurious currents in static and
moving droplets, and the propagation of damped capillary waves. 
Each of these tests were conducted for density contrasts spanning several orders of magnitude, 
including a realistic air-water configuration for the case of capillary waves. 
Both of our consistent methods demonstrate slight to moderate levels of increased accuracy
when compared to the standard version of our method that is not 
consistent with respect to mass-momentum transport. 
A second-order error convergence rate is observed in the case of the 
static droplet and the capillary wave, with the rate
reverting to first-order in the case of droplet advection.
Finally, the robustness and stability of the present method was demonstrated
using the case of a raindrop falling in air, which involves complex interactions
of surface tension forces, viscosity and inertia. The method is able to deliver
relatively good estimates of the flow features especially at low to moderate resolutions, 
as evidenced by the smooth evolution of quantities such as the droplet kinetic
energy and moment of inertia. A key feature of the our class of consistent methods 
is the ability to circumvent un-physical interfacial deformations (artificial atomization of the raindrop)
that are rampant in the case of the non consistent version of the method.
