\chapter*{Acknowledgement}
\addcontentsline{toc}{chapter}{Acknowledgement} % Add the preface to the table of contents as a chapter

The completion of the present body of work would not be possible without the perennial support, guidance and intellectual stimulation provided by my supervisors Prof. St\'ephane Zaleski and Dr. Daniel Fuster. Their everlasting patience, combined with their insights as experts in the field of mathematical and computational physics has made an immense contribution to my education as a budding scientist. I am indebted to Prof. Gaurav Tomar for providing me a gateway into the scientific discipline and academia, and Prof. St\'ephane Zaleski for recognizing my potential as a master's student and hence handing me the opportunity to carry out doctoral studies under his tutelage. I sincerely thank Dr. Wojciech Aniszewski for his unconditional mentorship during the initial years of my thesis. In addition to my immediate supervisors, I am also grateful to Dr. St\'ephane Popinet for the numerous instructive discussions. The numerical investigations constituting the thesis would not be possible without the combined contributions of the numerous scientists involved in the development of the ``PARIS Simulator" and ``Basilisk" codes. I convey my heartfelt appreciation to Dr. Jean-Luc Estivalezes and Dr. Lydia Bourouiba for investing their time and effort into monitoring my progress, as well as their valuable feedback during the review process.  

I would also like to convey my gratitude towards Simona Otarasanu, Barend Bentvelsen, Augustin Guibaud, Serena Costanzo, as well as all of my present coworkers and former colleagues, without whom, life in the lab would not be quite so colorful. Throughout my time as a graduate student, I have had the incredible fortune of having a continuous source of emotional support and joy in the form of Samta Kothari, Gaurav Prabhudesai, Anastasia Loukrezi, Christina Agapopoulou, François Chantriaux and Georg Lopez.  
Last but not the least, I would like to thank my family : my parents and my late grandparents, whose unwavering belief in my capabilities and relentless encouragement have made this difficult journey possible.    

